GravExplain: Continuous gravitational wave searches in a table-top experiment

### 2020-02-06 paper/paper_main.tex
68th day of Summer

@@

% overview
In this article we describe how a Michelson interferometer can be used to demonstrate the search for continuous waves to a general audience. 
The techniques used are very similar to those used by researchers to search for continuous waves in LIGO and Virgo data. 
In section X we describe...... \han{add this in at the end}

% \section{Method and results}
\section{Table-top gravitational wave science}

Detectable gravitational waves are caused by the most violent events in the universe. Like two massive black-holes merging with each other. Even so, the effects they produce are on the order of one part in $10^{21}$ for our best detectors. Given the impossibility of such detection with a table-top experiment, we demonstrate the same techniques but for observing sound waves.

@@

A Michelson interferometer is a well-known optics configuration used in many undergraduate and research labs, including in the detectors for gravitational waves. A laser beam of coherent (monochromatic and equal phase) light is split into two arms at right-angles to each other, travelling down and back from a mirror at the end of each arm. The beams return to interfere at a screen and produce a pattern of concentric, circular fringes.

\jam{See appendix for the exact set-up, lenses are included to focus the beam and view the interference pattern ||| include Michelson schematic with speaker, webcam, and photodiode}.

@@

\begin{figure}
	\includegraphics[width=\textwidth]{webcam_still0.pdf}
	\caption{Interference pattern from Michelson interferometer as viewed by webcam from below (hence why the pattern appears non-circular), red dot indicates point where intensity time series was measured \jam{do I need to add a scale to this picture?}}
	\label{fig:interference_pattern}
\end{figure}

@@

\subsection{The webcam method}
% how to analyse video, python OpenCV2

@@

@@

\subsection{Tracking tones with the Viterbi algorithm}
% explaination and successful application
% lean on the continuous gravitational waves angle

@@

@@

To demonstrate this technique, we inject a tone that wanders in frequency over time, remembering to integrate up the phase change over time (see Appendix~\ref{app:phase_gotcha} for why). Alternatively, a function generator can be slowly controlled to achieve the same input. Shown in Figure~\ref{fig:viterbi_overlay} is the resultant spectrogram of the observed signal, along with the Viterbi path and the injected signal. With great success, the Viterbi algorithm recovers the injected frequency over time, likely due to the high signal-to-noise ratio (SNR). Note that the signal initially moves too quickly (more than 0.11 Hz per second) for Viterbi to track, and that there appears to be an anomaly at 150s, likely a disturbance around the interferometer. Decreasing the signal-to-noise ratio closer to unity or increasing the allowed change in frequency over time causes Viterbi to more easily lose the signal. However, the demonstration clearly shows the ability to recover a continuous wave with the Viterbi algorithm, just as is done in continuous gravitational wave searches.

@@

\jam{What more is there to say about Viterbi?}

\section{Optical microphone}
% aim of optical microphone: play back injected sounds

@@

The entire circuit diagram is detailed and shown in Appendix~\ref{app:circuit_diagram}.

\subsection{Initial recordings}
% breakdown of all the filters used along the way

The optical microphone was tested on recordings of one human talking and on music ranging from simple melodies to full tracks. Each recording was around one minute long with processing done on ten second segments. For each observation, the audio was played through the speaker with minimal activity around interferometer as the Pi script recorded the signal. The saved time series was then directly converted to a .wav file and played as an audio recording.

@@

@@

Applying an existing implementation of logMMSE to the optical microphone recordings produces dramatically cleaner results. The time series and spectrum from before and after the logMMSE filter is applied are shown in Figures~\ref{fig:logMMSE_timeseries}~\&~\ref{fig:logMMSE_spectrum}. All of the mains harmonics are significantly attenuated. And further high-passing removes the low frequency curve beneath 100Hz.

@@

\subsection{Future work}
% full model of transfer function
A full model of the transfer function would detail the system starting from the voltage sent to the speaker and ending with the value recorded by the Pi, including any non-linearities in the speaker audio production, the speaker-mirror coupling, the separation-intensity relation, and the photodiode reading. Although we have the separation-intensity relation as linear for small motions, the other components of this full model are yet to be determined.

@@

\section{Conclusions}
% we are able to successfully ...
Using a Michelson interferometer, we able to successfully demonstrate the Viterbi algorithm recovering an injected signal. Just as is done is continuous gravitational wave searches. We are also able to play-back chordal or rhythmic music from an optical microphone (using the same interferometer), while recordings of simple speech are audible but unintelligible.

\appendix

@@

\jam{do I need to include access to sample .mp4 and .wav recordings: source, raw, and processed?}
\jam{this should be neatened, make all .py and move all .ipynb in here}

@@

\section{Michelson interferometer schematic}
\jam{please include}

Note that we do not include the common compensating plate that equalises the amount of time spent in glass for each beam. We did not find it necessary to include it to be able to demonstrate the results desired.

@@

\begin{figure}%[H]
	\includegraphics[width=\textwidth]{circuit_diagram.pdf}
	\caption{Circuit diagram for photodiode reading. Photodiode in reverse-bias over an op-amp, analog signal then passed through Sallen-Key anti-aliasing filter tuned to 16kHz, then into an analog-to-digital converter (ADC), that’s finally read by the special purpose input (SPI) pins of a Raspberry Pi}
	\label{fig:circuit_diagram}
\end{figure}

>>>>

% cover page
\section*{Project contributions}
% contributions clarify what’s to be assessed
Hannah Middleton: introduction of paper, design and making of interferometer

James W. Gardner: all other paper sections, design and making of photodiode circuit, and all of the source code

\jam{do we need to repeat those people in the Acknowledgements?}

\newpage

@@

% overview
In this article we describe how a Michelson interferometer can be used to demonstrate the search for continuous gravitational waves to a general audience. 
The techniques used are very similar to those used by researchers to search for continuous waves in LIGO and Virgo data.
In Section~\ref{sec:ifo} we detail the interferometer set-up, Section~\ref{sec:single_tone} demonstrates observing a single tone, Section~\ref{sec:viterbi_wandering} details tracking a wandering frequency tone, and Section~\ref{sec:optical_microphone} extends the set-up to be able to capture complex audio, such as music and speech.

% \section{Method and results}
\section{Table-top gravitational wave science}
\label{sec:ifo}

Detectable gravitational waves are caused by the most violent events in the universe. Like two massive black-holes merging with each other. Even so, the effects they produce are on the order of one part in $10^{21}$ for our best detectors\cite{GW150914}. Given the impossibility of such detection with a table-top experiment, we demonstrate the same techniques but for observing sound waves.

@@

A Michelson interferometer is a well-known optics configuration used in many undergraduate and research labs, including in the detectors for gravitational waves. A laser beam of coherent (monochromatic and equal phase) light is split into two arms at right-angles to each other, travelling down and back from a mirror at the end of each arm. The beams return to interfere at a screen and produce a pattern of concentric, circular fringes. The configuration is shown in Figure~\ref{fig:ifo_schematic_webcam}.

\begin{figure}
	\includegraphics[width=0.8\textwidth]{ifo_schematic_webcam.pdf}
	\caption{Michelson interferometer schematic, interference pattern is recorded by a webcam}
	\label{fig:ifo_schematic_webcam}
\end{figure}
% Note that we do not include the common compensating plate that equalises the amount of time spent in glass for each beam. We did not find it necessary to include it to be able to demonstrate the results desired.

@@

\begin{figure}
	\includegraphics[width=\textwidth]{webcam_still0.pdf}
	\caption{Interference pattern from Michelson interferometer as viewed by webcam from below (hence why the pattern appears non-circular), red dot indicates point where intensity time series was measured, the central fringe is approximately 5mm across}
	\label{fig:interference_pattern}
\end{figure}

@@

\section{Observing a single tone}
\label{sec:single_tone}
% the webcam method
% how to analyse video, python OpenCV2

\subsection{Method}

@@

\subsection{Results}

@@

\section{Observing a wandering tone}
\label{sec:viterbi_wandering}
% Tracking tones with the Viterbi algorithm
% explaination and successful application
% lean on the continuous gravitational waves angle

\jam{need to cite the various searches here}

@@

\subsection{The Viterbi algorithm}

@@

\subsection{Results}

To demonstrate this technique, we inject a tone that wanders in frequency over time, remembering to integrate up the phase change over time (see Appendix~\ref{app:phase_gotcha} for why). Alternatively, a function generator can be slowly controlled to achieve the same input. Shown in Figure~\ref{fig:viterbi_overlay} is the resultant spectrogram of the observed signal, along with the Viterbi path and the injected signal.
To the eye, the Viterbi algorithm recovers the injected frequency over time, likely due to the high signal-to-noise ratio (SNR). Quantifying this, the Viterbi path stays within one frequency bin (approximately 0.3Hz) of the injected tone for 94\% of the run. Note that the signal initially moves too quickly (more than 0.11 Hz per second) for Viterbi to track, and that there appears to be an anomaly at 150s, likely a disturbance around the interferometer.
Decreasing the signal-to-noise ratio closer to unity or increasing the allowed change in frequency over time causes Viterbi to more easily lose the signal. However, the demonstration clearly shows the ability to recover a continuous wave with the Viterbi algorithm, just as is done in continuous gravitational wave searches.

@@

\section{Complex audio}
\label{sec:optical_microphone}
% Optical microphone
% aim of optical microphone: play back injected sounds

@@

The entire circuit diagram is detailed and shown in Appendix~\ref{app:circuit_diagram}. The updated schematic is shown in Figure~\ref{fig:ifo_schematic_podo}.

\begin{figure}
	\includegraphics[width=0.8\textwidth]{ifo_schematic_photodiode.pdf}
	\caption{Michelson interferometer schematic, a part of the interference pattern is recorded by a photodiode}
	\label{fig:ifo_schematic_podo}
\end{figure}

\subsection{Initial results}
% breakdown of all the filters used along the way

The optical microphone was tested on recordings of one human talking and on music ranging from simple melodies to full tracks. Each recording was around one minute long with processing done on ten second segments. For each observation, the audio was played through the speaker with minimal activity around interferometer as the Pi script recorded the signal. The saved time series was then directly converted to a .wav file and played as an audio recording, using the \textbf{scipy.io.wavfile.write}\cite{scipy} implementation in Python\cite{python}.

@@

\subsection{Filters applied}

@@

Applying an existing implementation of logMMSE\cite{logmmse} to the optical microphone recordings produces dramatically cleaner results. The time series and spectrum from before and after the logMMSE filter is applied are shown in Figures~\ref{fig:logMMSE_timeseries}~\&~\ref{fig:logMMSE_spectrum}. All of the mains harmonics are significantly attenuated. And further high-passing removes the low frequency curve beneath 100Hz.

@@

\section{Future work}

A full model of the transfer function would detail the system starting from the voltage sent to the speaker and ending with the value recorded by the Pi, including any non-linearities in the speaker audio production, the speaker-mirror coupling, the separation-intensity relation, and the photodiode reading. Although we have the separation-intensity relation as linear for small motions, the other components of this full model are yet to be determined.

@@

\jam{should this be more broad a future work section?}

\section{Conclusions}

Using a Michelson interferometer, we able to successfully demonstrate the Viterbi algorithm recovering an injected signal. Just as is done is continuous gravitational wave searches. We are also able to play-back chordal or rhythmic music from an optical microphone (using the same interferometer), while recordings of simple speech are audible but unintelligible.

\newpage
\appendix

@@

\jam{update this to the current README.txt}

This project is implemented in python3\cite{python} inside of jupyter notebooks\cite{jupyter}\cite{ipython} and uses the packages of numpy\cite{numpy}, scipy\cite{scipy}, matplotlib\cite{matplotlib}, tqdm\cite{tqdm}, and logmmse\cite{logmmse}.

@@

@@

\begin{figure}%[H]
	\includegraphics[width=\textwidth]{circuit_diagram_2.pdf}
	\caption{Circuit diagram for photodiode reading. Photodiode in reverse-bias over an op-amp, analog signal then passed through Sallen-Key anti-aliasing filter tuned to 16kHz, then into an analog-to-digital converter (ADC), that’s finally read by the special purpose input (SPI) pins of a Raspberry Pi}
	\label{fig:circuit_diagram}
\end{figure}
